Partial Least Squares Regression with Multiple Functional and Scalar Predictors

### 2025-08-12 manuscript/main.tex
regularized estimation notation improved

\begin{equation}\label{def:squared_empirical_cov}
	\widehat{\textnormal{Cov}} 
	( \langle 
	\widetilde{W},
	\xi \rangle_{\mathcal{H}}, Y )
	:=
	\frac{1}{n} 
	\sum_{i=1}^n
	y_i
	\langle \widetilde{W}_i, \xi \rangle_{\mathcal{H}}
	.
\end{equation}
We denote the estimated PLS component direction as
\begin{equation}\label{def:maximizer_squared_empirical_cov}
	\hat{\xi} 
	:= \arg \max_{\xi \in \widetilde{\mathcal{H}}
	}~\widehat{\textnormal{Cov}}^2 
	\bigl(
	\langle \widetilde{W}, \xi \rangle_{\mathcal{H}}, Y 
	\bigr)~s.t.~\| \xi  \|_\mathcal{H} = 1.
\end{equation}
%
Here, $\hat{\xi} \in \widetilde{\mathcal{H}}$ is an ordered pair which can be expanded as:
\begin{equation*}
	\hat{\xi}	= 
	\bigl(
	\hat{\xi}_1(t), \ldots, \hat{\xi}_K(t), \hat{\boldsymbol{\zeta}}
	\bigr)
	=
\biggl(
	\sum_{m=1}^M \hat{\gamma}_{1m} \, b_m(t),
		 \ldots, 
		\sum_{m=1}^M \hat{\gamma}_{Km} \, b_m(t), \hat{\boldsymbol{\zeta}}
	\biggr)
	,
\end{equation*}
where  $\hat{\boldsymbol{\zeta}}$ is the scalar part. Obtaining these coefficients is equivalent to solving the maximization problem   \eqref{def:maximizer_squared_empirical_cov}. The following proposition formulates this coefficients obtaining procedure as a generalized Rayleigh quotient:
%
\begin{proposition}\label{proposition:eigen_noregul}
	Let
	$
	(\mathbb{B}, \mathbb{B}^{\prime \prime}, \Theta, \mathbf{y})
	$
	denote the observed data defined in \eqref{def:problem_instance}.
	At the $l$-th iteration of the PLS algorithm, 
	the coefficients of the squared covariance  maximizer defined in \eqref{def:maximizer_squared_empirical_cov}, 
	is obtained as
	\begin{equation}	\label{eq: Unregularized generalized rayleigh quotient equation}
		\left(
		\hat{\gamma}_{11}, \ldots, \hat{\gamma}_{1M}
		, \ldots,
		\hat{\gamma}_{K1}, \ldots, \hat{\gamma}_{KM}
		, \hat{\boldsymbol{\zeta}}^\top
		\right)^\top
		=
		\arg \max_{\boldsymbol{\xi} \in \mathbb{R}^{MK+p}}   \boldsymbol{\xi}^\top V \boldsymbol{\xi}
		\quad \text{subject to} \quad \boldsymbol{\xi}^\top \mathbb{B} \boldsymbol{\xi} = 1.
	\end{equation}
	where
	\begin{equation}\label{def:V_matrix}
		V := \frac{1}{n^2} (\mathbb{B} \Theta^\top \mathbf{y})(\mathbb{B} \Theta^\top \mathbf{y})^\top \in \mathbb{R}^{(MK+p) \times (MK+p)}.
	\end{equation}
\end{proposition}
The proof of Proposition \ref{proposition:eigen_noregul}
is provided in Appendix \ref{section:proof:proposition:linear_regul_system_unregularized}.

@@

To avoid the orthogonal basis assumption of  \citet{Rice1991}, we adopt the strategy of \citet{Silverman1996}, which replaces the standard orthonormality constraint with a weaker one based on a modified inner product that incorporates roughness. Accordingly, our estimation procedure at the $l$-th iteration, iteration index omitted and assuming the observations have been residualized in previous steps, solves the following optimization problem:
\begin{equation}\label{def:maximizer_squared_empirical_cov_reg}
	\hat{\xi} 
	= 
	(\hat{\xi}_1, \ldots, \hat{\xi}_K, \hat{\boldsymbol{\zeta}}
	)
	:= \arg \max_{\xi \in \widetilde{\mathcal{H}}
	}~\widehat{\textnormal{Cov}}^2 
	\bigl(
	\langle \widetilde{W}, \xi \rangle_{\mathcal{H}}, Y 
	\bigr)~s.t.~\| \xi  \|_\mathcal{H} +  \sum \limits_{j=1}^K \lambda_j 
	\operatorname{PEN}(\xi_K) = 1.
\end{equation}
This formulation maximizes the squared covariance over a class of smooth functions.
Proposition~\ref{proposition:eigen_regul} expresses this problem as a generalized Rayleigh quotient.

\begin{proposition}\label{proposition:eigen_regul}
	Let
	$
	(\mathbb{B}, \mathbb{B}^{\prime \prime}, \Theta, \mathbf{y})
	$
	denote the observed data defined in \eqref{def:problem_instance}.
	At the $l$-th iteration of the PLS algorithm, the maximizer of  the squared empirical covariance defined in \eqref{def:maximizer_squared_empirical_cov_reg}, represented as the Euclidean vector form as in \eqref{def:maximizer_euclidean_form},
	is obtained by solving 
	\begin{equation}	\label{eq: Regularized generalized rayleigh quotient equation}
		\left(
		\hat{\boldsymbol{\gamma}}_1^\top, \ldots, \hat{\boldsymbol{\gamma}}_K^\top, \hat{\boldsymbol{\zeta}}^\top
		\right)^\top
		=
		\arg \max_{\boldsymbol{\xi} \in \mathbb{R}^{MK+p}}   \boldsymbol{\xi}^\top V \boldsymbol{\xi}
		\quad \text{s.t.} \quad 
		%
		% constraint
		\boldsymbol{\xi}^\top (\mathbb{B} + \Lambda \mathbb{B}^{\prime \prime}) \boldsymbol{\xi} = 1.
	\end{equation}
	where we recall that $
	V = n^{-2} (\mathbb{B} \Theta^\top \mathbf{y})(\mathbb{B} \Theta^\top \mathbf{y})^\top$ is defined in  \eqref{def:V_matrix}, and 
	$\Lambda \in \mathbb{R}^{(MK+p) \times (MK+p)}$ is given by
	\begin{equation}\label{def:Lambda}
		\Lambda := \operatorname{blkdiag}(\lambda_1 I_M, \ldots, \lambda_K I_M, 0_{p \times p}),
	\end{equation}
	where $0_{p \times p}$ denotes the $p \times p$ zero matrix.
\end{proposition}
The proof of Proposition \ref{proposition:eigen_regul} is provided in Appendix  \ref{section:proof:proposition:eigen_regul}.
%orthonormality preview
The constraint $	\boldsymbol{\xi}^\top (\mathbb{B} + \Lambda \mathbb{B}^{\prime \prime}) \boldsymbol{\xi} = 1$
enforces the orthonormality of the estimated PLS component directions with respect to a modified inner product (see Section \ref{section:sub:prop_geom} for details).
The smoothing parameter $\lambda_k > 0$  balances goodness of fit and smoothness in $\hat{\xi}_j$. Smaller $\lambda_k$ yields components that better fit the data but risks overfitting; setting $\lambda_k = 0$ recovers the unregularized solution in Proposition~\ref{proposition:eigen_noregul}. Larger $\lambda_k$ enforces greater smoothness, and in the limit $\lambda_k \to \infty$, $\hat{\xi}_j(t)$ approaches a linear form $a + bt$. In practice, both $\{\lambda_k\}$ and the number of components $L$ can be selected via cross-validation using a predictive criterion such as mean squared error.

>>>>

\begin{equation}\label{def:squared_empirical_cov}
	\widehat{\textnormal{Cov}} 
	( \langle 
	\widetilde{W},
	\xi \rangle_{\mathcal{H}}, Y )
	:=
	\frac{1}{n} 
	\sum_{i=1}^n
	y_i
	\langle \widetilde{W}_i, \xi \rangle_{\mathcal{H}}
	.
\end{equation}
We denote the estimated PLS component direction as
\begin{equation}\label{def:maximizer_squared_empirical_cov}
	\hat{\xi} 
	:= \arg \max_{\xi \in \widetilde{\mathcal{H}}
	}~\widehat{\textnormal{Cov}}^2 
	\bigl(
	\langle \widetilde{W}, \xi \rangle_{\mathcal{H}}, Y 
	\bigr)~s.t.~\| \xi  \|_\mathcal{H} = 1.
\end{equation}
%
Here, $\hat{\xi} \in \widetilde{\mathcal{H}}$ is an ordered pair  expanded as:
\begin{equation*}
	\hat{\xi}	= 
	\bigl(
	\hat{\xi}_1(t), \ldots, \hat{\xi}_K(t), \hat{\boldsymbol{\zeta}}
	\bigr)
	=
\biggl(
	\sum_{m=1}^M \hat{\gamma}_{1m} \, b_m(t),
		 \ldots, 
		\sum_{m=1}^M \hat{\gamma}_{Km} \, b_m(t), \hat{\boldsymbol{\zeta}}
	\biggr)
	,
\end{equation*}
where  $\hat{\boldsymbol{\zeta}}$ is the scalar part. Obtaining these coefficients is equivalent to solving the maximization problem   \eqref{def:maximizer_squared_empirical_cov}. The following proposition formulates this coefficients obtaining procedure as a generalized Rayleigh quotient:
%
\begin{proposition}\label{proposition:eigen_noregul}
	Let
	$
	(\mathbb{B}, \mathbb{B}^{\prime \prime}, \Theta, \mathbf{y})
	$
	denote the observed data defined in \eqref{def:problem_instance}.
	At the $l$-th iteration of the PLS algorithm, 
	the coefficients of the squared covariance  maximizer defined in \eqref{def:maximizer_squared_empirical_cov}, 
	is obtained as
	\begin{equation}	\label{eq: Unregularized generalized rayleigh quotient equation}
		\left(
		\hat{\gamma}_{11}, \ldots, \hat{\gamma}_{1M}
		, \ldots,
		\hat{\gamma}_{K1}, \ldots, \hat{\gamma}_{KM}
		, \hat{\boldsymbol{\zeta}}^\top
		\right)^\top
		=
		\arg \max_{\boldsymbol{\xi} \in \mathbb{R}^{MK+p}}   \boldsymbol{\xi}^\top V \boldsymbol{\xi}
		\quad \text{subject to} \quad \boldsymbol{\xi}^\top \mathbb{B} \boldsymbol{\xi} = 1.
	\end{equation}
	where
	\begin{equation}\label{def:V_matrix}
		V := \frac{1}{n^2} (\mathbb{B} \Theta^\top \mathbf{y})(\mathbb{B} \Theta^\top \mathbf{y})^\top \in \mathbb{R}^{(MK+p) \times (MK+p)}.
	\end{equation}
\end{proposition}
The proof of Proposition \ref{proposition:eigen_noregul}
is provided in Appendix \ref{section:proof:proposition:linear_regul_system_unregularized}.

@@

To avoid the orthogonal basis assumption of  \citet{Rice1991}, we adopt the strategy of \citet{Silverman1996}, which replaces the standard orthonormality constraint with a weaker one based on a modified inner product that incorporates roughness. Accordingly, our estimation procedure at the $l$-th iteration, iteration index omitted and assuming the observations have been residualized in previous steps, solves the following optimization problem:
\begin{equation}\label{def:maximizer_squared_empirical_cov_reg}
	\hat{\xi} 
	:= \arg \max_{\xi \in \widetilde{\mathcal{H}}
	}~\widehat{\textnormal{Cov}}^2 
	\bigl(
	\langle \widetilde{W}, \xi \rangle_{\mathcal{H}}, Y 
	\bigr)~s.t.~\| \xi  \|_\mathcal{H} +  \sum \limits_{j=1}^K \lambda_j 
	\operatorname{PEN}(\xi_K) = 1.
\end{equation}
Here, $\hat{\xi} \in \widetilde{\mathcal{H}}$ is an ordered pair  expanded as:
\begin{equation*}
	\hat{\xi}	= 
	\bigl(
	\hat{\xi}_1(t), \ldots, \hat{\xi}_K(t), \hat{\boldsymbol{\zeta}}
	\bigr)
	=
	\biggl(
	\sum_{m=1}^M \hat{\gamma}_{1m} \, b_m(t),
	\ldots, 
	\sum_{m=1}^M \hat{\gamma}_{Km} \, b_m(t), \hat{\boldsymbol{\zeta}}
	\biggr)
	,
\end{equation*}
where  $\hat{\boldsymbol{\zeta}}$ is the scalar part.
This formulation maximizes the squared covariance over a class of smooth functions.
 Obtaining these coefficients is equivalent to solving the maximization problem   \eqref{def:maximizer_squared_empirical_cov_reg}. The following proposition formulates this coefficients obtaining procedure as a generalized Rayleigh quotient:
\begin{proposition}\label{proposition:eigen_regul}
	Let
	$
	(\mathbb{B}, \mathbb{B}^{\prime \prime}, \Theta, \mathbf{y})
	$
	denote the observed data defined in \eqref{def:problem_instance}.
	At the $l$-th iteration of the PLS algorithm, 	the coefficients of the squared covariance  maximizer defined in \eqref{def:maximizer_squared_empirical_cov_reg}, 
	is obtained as
	\begin{equation}	\label{eq: Regularized generalized rayleigh quotient equation}
	\left(
\hat{\gamma}_{11}, \ldots, \hat{\gamma}_{1M}
, \ldots,
\hat{\gamma}_{K1}, \ldots, \hat{\gamma}_{KM}
, \hat{\boldsymbol{\zeta}}^\top
\right)^\top
		=
		\arg \max_{\boldsymbol{\xi} \in \mathbb{R}^{MK+p}}   \boldsymbol{\xi}^\top V \boldsymbol{\xi}
		\quad \text{s.t.} \quad 
		%
		% constraint
		\boldsymbol{\xi}^\top (\mathbb{B} + \Lambda \mathbb{B}^{\prime \prime}) \boldsymbol{\xi} = 1.
	\end{equation}
	where we recall that $
	V = n^{-2} (\mathbb{B} \Theta^\top \mathbf{y})(\mathbb{B} \Theta^\top \mathbf{y})^\top$ is defined in  \eqref{def:V_matrix}, and 
	$\Lambda \in \mathbb{R}^{(MK+p) \times (MK+p)}$ is given by
	\begin{equation}\label{def:Lambda}
		\Lambda := \operatorname{blkdiag}(\lambda_1 I_M, \ldots, \lambda_K I_M, 0_{p \times p}),
	\end{equation}
	where $0_{p \times p}$ denotes the $p \times p$ zero matrix.
\end{proposition}
The proof of Proposition \ref{proposition:eigen_regul} is provided in Appendix  \ref{section:proof:proposition:eigen_regul}.
%orthonormality preview
The constraint $	\boldsymbol{\xi}^\top (\mathbb{B} + \Lambda \mathbb{B}^{\prime \prime}) \boldsymbol{\xi} = 1$
enforces the orthonormality of the estimated PLS component directions with respect to a modified inner product (see Section \ref{section:sub:prop_geom} for details).
The smoothing parameter $\lambda_k > 0$  balances goodness of fit and smoothness in $\hat{\xi}_j$. Smaller $\lambda_k$ yields components that better fit the data but risks overfitting; setting $\lambda_k = 0$ recovers the unregularized solution in Proposition~\ref{proposition:eigen_noregul}. Larger $\lambda_k$ enforces greater smoothness, and in the limit $\lambda_k \to \infty$, $\hat{\xi}_j(t)$ approaches a linear form $a + bt$. In practice, both $\{\lambda_k\}$ and the number of components $L$ can be selected via cross-validation using a predictive criterion such as mean squared error.
